The \dealii Library, Version 9.1

### 2019-06-02 9.1/paper.tex
Write section on symbolic differentiation

The major changes of this release are:
%
\begin{itemize}
  \item Improved support for automatic differentiation (see
    Section~\ref{subsec:ad}),
  \item Full support for $hp$ adaptivity in parallel computations (see
    Section~\ref{subsec:hp}),
  \item Interfaces to the HDF5 file format and libraries (see
    Section~\ref{subsec:hdf5}),
  \item Significantly extended GPU support (see Section~\ref{subsec:gpu}),
  \item Parallel geometric multigrid (GMG) improvements (see
    \cite{ClevengerHeisterKanschatKronbichler2019} and
    Section~\ref{subsec:gmg}),
  \item Four new tutorial programs (step-61, step-62, step-63, step-64),
    as well as one new code gallery program (see
    Section~\ref{subsec:steps}).
\end{itemize}
%
The major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
\item
  The release contains bug fixes and performance improvements of the
  matrix-free framework and related geometric multigrid solvers. In
  particular, the implementation of the Chebyshev iteration, an often used
  smoother in the matrix-free context, has been revised to reduce vector
  accesses. This speeds up matrix-free multigrid solvers by up to 10-15\%
  on geometries with affine (parallelogram and parallelpiped) cells, and up
  to 5\% on geometries with cells bounded by curved edges and faces.

@@

>>>>

The major changes of this release are:
%
\begin{itemize}
  \item Improved support for automatic differentiation (see
    Section~\ref{subsec:ad}),
  \item Dedicated support for symbolic algebra (see
    Section~\ref{subsec:sd}),
  \item Full support for $hp$ adaptivity in parallel computations (see
    Section~\ref{subsec:hp}),
  \item Interfaces to the HDF5 file format and libraries (see
    Section~\ref{subsec:hdf5}),
  \item Significantly extended GPU support (see Section~\ref{subsec:gpu}),
  \item Parallel geometric multigrid (GMG) improvements (see
    \cite{ClevengerHeisterKanschatKronbichler2019} and
    Section~\ref{subsec:gmg}),
  \item Four new tutorial programs (step-61, step-62, step-63, step-64),
    as well as one new code gallery program (see
    Section~\ref{subsec:steps}).
\end{itemize}
%
The major changes are discussed in detail in Section~\ref{sec:major}. There
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
\item
  The release contains bug fixes and performance improvements of the
  matrix-free framework and related geometric multigrid solvers. In
  particular, the implementation of the Chebyshev iteration, an often used
  smoother in the matrix-free context, has been revised to reduce vector
  accesses. This speeds up matrix-free multigrid solvers by up to 10-15\%
  on geometries with affine (parallelogram and parallelpiped) cells, and up
  to 5\% on geometries with cells bounded by curved edges and faces.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dedicated support for symbolic algebra, including algebra differentiation}
\label{subsec:sd}
To complement the automatic differentiation features in \dealii{}, this release sees
the first step towards integrating and supporting a highly performant computer algebra 
system (CAS) via the SymEngine library.
The \texttt{Expression} class in the namespace \texttt{Differentiation::SD} interfaces
to SymEngine and forms the basis of symbolic computations, offering a full set of 
overloaded operators and a C++ style interface. This class offers the following 
basic functionality:
\begin{itemize}
\item symbolic variable definition,
\item symbolic function definition,
\item expression creation using standard C++ syntax,
\item expression parsing,
\item comparison operations,
\item logical operations,
\item conditional expression construction,
\item differentiation,
\item substitution (partial and complete), and
\item serialization.
\end{itemize}
We also provide an extensive set of math operations, with a syntax mimicing that used 
in the C++ standard library. Using \dealii{}'s \texttt{Expression} class as a basis, we 
have developed a set of functions that can be used to create \dealii{} \texttt{Tensor}s
and \texttt{SymmetricTensor}s of symbolic variables and symbolic functions. This
gives full symbolic tensor algebra support using the pre-existing \texttt{Tensor}
and \texttt{SymmetricTensor} class and associated functions. A set of utility functions to
\begin{itemize}
\item \texttt{differentiate} scalar expressions, tensor and symmetric tensors of expressions
with respect to other scalar expressions, tensor and symmetric tensors of expressions,
\item create symbolic substitution maps,
\item resolve explicit dependencies between expressions, and
\item perform scalar and tensor valued substitution (including conversion from symbolic to 
real-valued scalars and tensors)
\end{itemize}
have also been implemented.

In the next release we expect to implement classes to assist in performing assembly operations 
in the same spirit as that which has been done in the \texttt{Differentiation::AD} namespace,
although in a fully symbolic manner.
We will also address increase the performance of the \texttt{Expression} class by leveraging
the optimization capabilities of SymEngine, including common subexpression elimination (CSE), 
as well as by generating high performance code-paths to evaluate these expressions through the
use of a custom-generated \texttt{std::function} (so-called ``lambda'' optimization) or by 
compiling expressions using the LLVM JIT compiler.
